---
title: 'Details: Model goodness of fit in GAMLj'
nickname: details_goodness
bibliography: 
     - bib.bib
     - bib.yaml
link-citations: true
topic: all
category: details
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(fontawesome)
options(scipen=999)

```



```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```

`r keywords("jamovi, GLM, Mixed, effect size indices, R-squared, loglikelihood ratio test")`

`r version("3.0.0")` 

# Introduction

Here you can find information regarding the R-squared computed for different models and their associated inferential tests.

# GLM (linear model)

The R-squared produced by `r modulename()` is the standard R-squared produced by virtually any statistical software . It is computed as $SS_{model}$ divided by $SS_{total}$. In `r modulename()`, it is extracted from the R model object by `performance::r2(model)` (@parameters). The adjusted R squared is computed as 

$$R_{adj}^2={{SS_{model}-SS_{res} \cdot ({df_{model}/df_{res}) \cdot }}\over{ SS_{model}+SS_{res}(df_{res}+1)/df_{res}}}$$

The F-tests associated to the R-squares tests the null-hypothesis that the model predictors explain no variance, which is equivalent to compare the model at hand with an intercept-only model. 


# GzLM (Generalized linear Models)

For the generalized linear model, the R-squared


`r issues()`



# Additional references