---
title: "Mixed Models module technical details"
topic: mixed
category: details
nickname: mixed_specs
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false

---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```


# Degrees of freedom

As far as it is possible, inferential tests are based on an automatic selection of the degrees of freedom for the t-tests and the F-tests. 

* For the t-tests the module relies on the Satterthwaite approximation of degrees of freedom as it is implemented by the [lmerTest package](https://cran.r-project.org/web/packages/lmerTest/index.html). `Lmertest::summary()` produces t-test DF and p-values in the majority of cases. When the model does not converge, or some problem with the model occurs, the DF cannot be estimated and thus the p-values are not reported in the parameter estimates table. A warning note is displayed in those cases. See details in [lmerTest package](https://cran.r-project.org/web/packages/lmerTest/index.html).

* For the F-tests of the main model ("Fixed Effects ANOVA"), the module relies again on the Satterthwaite approximation of degrees of freedom as it is implemented by the [lmerTest package](https://cran.r-project.org/web/packages/lmerTest/index.html). `Lmertest::anova()` has difficulties in calculating F-test DF and p-values when the model does not converge, there are problems with the model random variances, or when numerical independent variables appear in complex interactions. When the `lmerTest::anova()`  does not produce the DF and p-values, the module switches to `car::Anova(..,type=3, test="F")` function of the R [car package](https://cran.r-project.org/web/packages/car/index.html). `car::Anova(..,type=3, test="F")` implements the Kenward-Roger method for degrees of freedom. A warning note is displayed in those cases.

* For F-tests of simple effects, the `car::Anova(..,type=3, test="F")` is always used, thus the the Kenward-Roger method for degrees of freedom is employed. This is because the `lmertest::anova()` requires the factor contrast to be `contr.sum()`, which does not allow to estimate simple effects. `car::Anova(..,type=3, test="F")` has not such a limitations, so it works fine for the purpose. 

```{r child="posthoc_specs.Rmd"}
```

# Cross-classifications

Assume one has schools and classes, coded as follows:

`r pic("commons/pics/compute_unique_clusters_0.png")`

We want to use school as a clustering variable, but the way `Class` is coded would pool together cases of different classes, because class 1, for instance, appears in all three schools, even if they are actually three different classes. 

To overcome this problem, some software allows specifying the clustering variable as `School:Class`, so classes can be distinguished. `r modulename()` does not allow that, because there's really no reason to it and it may be confusing. The solution in `r modulename()` is to recode the classes such that they all have a unique code. In `r jamovi` one can simply compute a new variable as follows:

`r pic("commons/pics/compute_unique_clusters_1.png")`

and obtains unique codes for the classes.

`r pic("commons/pics/compute_unique_clusters_2.png")`

These unique codes can now be used in mixed models to define class as unique clusters in the clustering variable.
`r backto("mixed","gmixed")`

