---
title: "Rosetta store: Nested like in R"
topic: contrast
category: rosetta
nickname: ros_mixed_nested
output: 
  html_document:
    
    includes:
         in_header: ganalytics.txt
    toc: true
    toc_float:
         collapsed: true
editor_options: 
  chunk_output_type: console
---


```{r setup, echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
data<-read.csv(mcdocs_path("data/toyclusters.csv"))

```

`r keywords("jamovi, R, Mixed Models, Multilevel Models, Nested design, Nested levels")`

`r version("1.0.0")` 

# Issue

_I have nested clustering variables (nested levels) and I am not sure the model is considering the nesting correctly._

# Short answer

Be sure that every cluster, at each level, has a unique value (factor level). If so, the model considers the nesting correctly

# Long answer

  
## Introduction

Mixed models are often used to analyze multilevel designs, in which several clustering variables appear. When more than one clustering variable exists, clusters can be nested or cross-classified. See `r mcdocs::link_pages(nickname="subjects_by_stimuli")` for some examples. In case of nested models, users sometimes wonder how to signal to `r modulename()` that the clustering variables are nested, meaning that they define different levels of a multilevel design. This uncertainty is often supported by the fact that `lmer()` command of `R package lme4` seems to require specifying the nested nature of the clustering variables with special notation (see below). However, this is not the case. Nested clustering can be specified both in `lmer()` command than in `r modulename()` without any special notation (model formula notation), and the models are correctly estimated taking into the account the nested nature of the design. Let's see a toy example to clarifying the issue.

## A quick three-level example

Assume you have three schools, within each school three classes, and pupils within classes. Schools are coded in variable `school` as `A`, `B`, and `C`. Assume you coded, within each school, the three classes as `1`, `2`, and `3`. Your dataset will look like this.
`r mcdocs::pic("rosetta/mixed_nested/coding1.png")`

This means that the designs features 3 schools, and 9 classes. Usually, you want to estimate the random coefficients (in this simple case only the intercepts) across schools and classes. If you run a model in `r modulename()` using these data, however, the model will pool together all pupils in class `1`, all pupils in class `2` an so on, irrespective of their school, as if pupils of different schools belonged to the same class. Indeed, in `r modulename()` you would have 3 schools and 3 classes :

`r mcdocs::pic("rosetta/mixed_nested/input1.png")`
`r mcdocs::pic("rosetta/mixed_nested/output1.png")`

This is equivalent to run in `R lme4` command

```{r, warning=FALSE,message=FALSE}
library(lmerTest)

data$school<-factor(data$school)
data$class<-factor(data$class)
lmer(y~1+(1|school)+(1|class),data=data)

```

## Solution

To avoid this issue, `R lmer()` allows using a different notation to specify the nesting. Namely

```{r, warning=FALSE,message=FALSE}
lmer(y~1+(1|school/class),data=data)

```

As you can see, the model has now 9 classes and 3 schools, as intended. However, the only thing that this notation does is to build a new variable `class:school` which assigns unique values to each class. Indeed, we can do that in R and in `r modulename()` to obtain exactly the same results. 

In R:

```{r, warning=FALSE,message=FALSE}
data$unique_class<-apply(cbind(data$class,data$school),1,paste,collapse="_")
data$unique_class<-factor(data$unique_class)

lmer(y~1+(1|school)+(1|unique_class),data=data)

```

In `r jamovi`:

`r pic("rosetta/mixed_nested/compute1.png")`
`r pic("rosetta/mixed_nested/input2.png")`
`r pic("rosetta/mixed_nested/output2.png")`

Indeed, the new variable `unique_class` is simply assigning unique values to the classes. 

`r pic("rosetta/mixed_nested/coding2.png")`

# Recap

When dealing with nested levels, be sure that each group (cluster) is coded with a unique code, and the nesting structure will be automatically taken into the account by the model.


# Rosetta's files

`r list_pages(category="rosetta")`


`r issues()`