---
title: "Simple Effects and Simple Interactions"
author: "Marcello Gallucci"
nickname: simple_effects
topic: glm
category: rosetta
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()

```


`r keywords("moderated regression, simple effects, simple interaction, linear model")`

`r version("3.0.0")`

In this page we work out some analysis involving simple effects and simple interaction, using jamovi `r modulename()`. We will replicate the analyses found in the very instructive webpage [Implementation of Three-factor Between-group Analysis of Variance](https://bcdudek.net/3wayanova/3way_factorial_Rmarkdown.htm) by **Bruce Dudek**. In particular, the analyses on [simple effects and interactions](https://bcdudek.net/3wayanova/3way_factorial_Rmarkdown.html#use-of-the-phia-package-for-contrasts-and-simple-effects).  We employ data from Keppel and Wickens, pg 466 data set, a 3x3x2 design  [keppel_3way_pg466.csv file](link to file). Dudek's page employes a variaty of R commands and packages to obtain the results, so we can compare `r modulename()` with different source of results. The aim of these analyses is to validate the results obtained with `r modulename()` with the ones obtained with different software.


# The research design

The dataset has a dependent variable, number of words recalled from a memorized list. Three independent variables defined as factors 1) _Grade_,  either fifth or twelfth graders 2) _Feedback_ conditions: control (none), praise, or negative. _WordType_, that is the type of words used: HF_HE, HF_LE, LF_LE



# GLM

## Input

We first set up a general linear model to replicate the overall ANOVA presented in [Dudek's page, par 3.1.2 ](https://bcdudek.net/3wayanova/3way_factorial_Rmarkdown.html#way-factorial-design.-a-completely-randomized-between-subjects-factorial-design).  We launch `General Linear Model` from the `Linear Models` menu and set up the model 

`r pic("examples/glm3/input.1.png")`

And we get (among other tables) the ANOVA results.

`r pic("examples/glm3/output.1.png")`

In R, we can use:

```{r}
data<-read.csv("../data/keppel_3way_pg466.csv",stringsAsFactors=T)
head(data)
data$Feedback <- car::recode(data$Feedback, "'none'='None'; 'pos'='Positive'; 'neg'='Negative'")
data$Feedback<-factor(data$Feedback,levels=c("None","Positive","Negative"))
data$WordType <- factor(data$WordType,
                    levels=c("LF_LE","HF_LE","HF_HE"))                  
contrasts(data$Grade)<-c(0,1)
model<-lm(numrecall~Feedback*WordType*Grade,data=data)
car::Anova(model,type=3)
emmeans::emmeans(model,specs=~Feedback*WordType)
```

# Contrasts

In Dudek's analysis, both _Feedback_ and _WordType_ are coded with custom orthogonal contrasts. The first contrast is a quadratic contrast (-1,-1,2) for _Feedback_ (2,-1,-1) for _WordType_. The second contrast is a linear contrast, (-1,-1,0) for _Feedback_ and (0,-1,1) for _WordType_. In other words, we are testing:

* Feedback:
    * first: None and Positive  vs Negative
    * second: None vs Negative

* WordType:
    * first: LF_LE  vs  HF_LE and HF_HE
    * second: HF_LE vs HF_HE

Both in R and `r modulename()` we employ polynomial contrasts that, opportunely re-arranged, yield the same results (same F-test, same t-test and p-values). Recall, in fact, that the scale and the sign of the contrasts are immaterial for the test results. In R, we can achieve this by changing the contrasts definition for the variables. For _Feedback_:

```{r}
contrasts(data$Feedback)<-contr.poly(3)
zapsmall(contrasts(data$Feedback))
```

However, the order of the contrasts codes, and thus the comparisons, is different (in Dudek's "Negative" as code 0 for the linear contrast and None has codes of the same sign). We can rearrange the coding as follows:

```{r}
poly<-contr.poly(3)
contrasts(data$Feedback)<- poly[c(3,2,1),]
zapsmall(contrasts(data$Feedback))
```

The same goes for _WordType_:

```{r}
contrasts(data$WordType)<-contr.poly(3)
zapsmall(contrasts(data$WordType))
```
In Dudek's, "LF_LE" is the zero-code level in linear, so we re-arrange:

```{r}
contrasts(data$WordType) <- poly[c(2,3,1),]
zapsmall(contrasts(data$WordType))

```

Finally, although this would not be recommended in standard analysis, the variable _Grade_ is coded with treatment code. This is not recommended because the contrasts coefficients for all factors are estimated for $Grade=0$, in the example "Grade 12th".

```{r}
(contrasts(data$Grad) <- c(1,0))

```

In `r modulename()`  we simply need to re-arrange the order of the factor levels in the datasheet to be sure that the polynomial contrasts are exactly what we want. In `r modulename()`, for 3-level factors, the polynomial contrasts are always: 

* Linear: first vs last level
* Quadratic: first and last versun second level

Thus, for need to order the levels as follows:

* Feedback:  Positive Negative None

* WordType: HF_HE LF_LE HF_LE

`r pic("examples/glm3/input.2.png")`

`r pic("examples/glm3/input.3.png")`

At this point, we can set the correct coding system for the factors

`r pic("examples/glm3/input.4.png")`

## Contrast coefficients

```{r }
model<-lm(numrecall~Feedback*WordType*Grade,data=data)
summary(model)
```

It can be verified that the results are identical to Dudek's results, with the exception of the sign and scale of the estimates, but all t-test and p-values are identical. The order of the effects in the table is different, but all terms are present.

Exactly the same results are obtained in `r modulename()`

`r pic("examples/glm3/output.2.png")`

Having established that the model results are the same, we need to code `Grade` with a centered coding system, so the results of the R `lm()` command are in line with the one expected from a standard ANOVA.

```{r }
contrasts(data$Grade)<-contr.sum(2)
model<-lm(numrecall~Feedback*WordType*Grade,data=data)
```

In `r modulename()`, we have 

`r pic("examples/glm3/input.7.png")`


# Simple interactions


Here we test the simple interaction _Feedback*WordType_ at different levels of grade (cf. Dudek's page, par 6.1). In R we can employ `testInteractions` command from `phia Package`.

```{r}

phia::testInteractions(model,fixed=c("Grade"), across=c("Feedback", "WordType"),adjust="none")


```


In `r modulename()` we ask for the `Simple Effects` of `Feedback` with `Moderators` both `WordType` and `Grade`, and select the option `Simple Interactions`.

`r pic("examples/glm3/input.5.png")`

The results show the F-test associated with the simple interactions at each level of _Grade_, and the individual interaction contrasts that inform the simple interactions.

`r pic("examples/glm3/output.5.png")`

Another possible set of simple interactions can be the interaction _WordType*Grade_ at different levels of _Feedback_. 

In R we have

```{r}
phia::testInteractions(model,fixed=c("Feedback"), across=c("Grade", "WordType"),adjust="none")

```

In `r modulename()`, we simply change the input of the `Simple Effect` options, putting _Feedback_ as the last moderator.

`r pic("examples/glm3/input.6.png")`

`r pic("examples/glm3/output.6.png")`

# Simple Effects

We now check that the results of Simple effects are the same R and in `r modulename()`. In Dudek's analysis, both simple effects with one moderator (named there _simple main effects_) and with more than one moderator (named there _simple-simple main effects_) are estimates. We can do both.

As regard simple effect with only one moderator, we can estimate, for instance, the simple effect of _Feedback_ at different levels of _WordType_. 

In R:

```{r}
phia::testInteractions(model,fixed=c("Feedback"), across=c("WordType"),adjust="none")

```

In `r modulename()`, we set the input of the `Simple Effect` option as required:

`r pic("examples/glm3/input.8.png")`

`r pic("examples/glm3/output.8.png")`

Again, the same results.

We now check that the results of Simple effects with more than one moderators.  For instance, one can estimate the effect of _WordType_ for all combinations of _Feedback_ and _Grade_ levels. 

In R:

```{r}

phia::testInteractions(model,fixed=c("Feedback","Grade"), across="WordType",adjust="none")

```

In `r modulename()`:

`r pic("examples/glm3/input.6.png")`

`r pic("examples/glm3/input.7.png")`

Obtaining the same results.



`r include_examples("glm")`

`r issues()`